---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analyzing Differences in Iowa Counties' Spending

#### Alexander M. Todd - November 23rd, 2025

## Introduction

Government's spend money on various different services. With many ways to collect and spend money, there is significant variety in how a Government allocated funds. The goal of this report is to understand the differences in how the various counties in Iowa collect, allocate, and spend its wealth.

The questions explored in this report are:

1. 

2. 

3. 

4. 

## Data

### Original Data

The data used in this report comes from the [State of Iowa Open Data Website](https://data.iowa.gov/). There were 5 sets of data utilized in this report:

1.  [Annual Personal Income by County](https://data.iowa.gov/Economic-Statistics/Annual-Personal-Income-for-State-of-Iowa-by-County/st2k-2ti2/about_data)

Included is the 'Personal Income' and 'Personal Income Per Capita' for each Iowa County for each year between 1997 and 2022.

2.  [County Actual Revenues](https://data.iowa.gov/Local-Government-Finance/County-Actual-Revenues-By-Type-By-Fiscal-Year/fp2u-f5hd/about_data)

Included is the Total Revenue gained by each Iowa County for each year between 2010 and 2024. There are many variables that split the Revenue by type, most notably the revenue levied by Property Tax and that provided by county issued goods and services.

3.  [County Budgeted Expenditures](https://data.iowa.gov/Local-Government-Finance/County-Budgeted-Expenditures-By-Service-Area-By-Fi/gk9s-gz9c/about_data)

Included is the budgeted spending for each Iowa County for each year between 2005 and 2026. The budget is split based on type of spending for which the money is allocated in the budget.

4.  [County Actual Expenditures](https://data.iowa.gov/Local-Government-Finance/County-Actual-Expenditures-by-Service-Area-by-Fisc/fxbr-vb9c/about_data)

Included is actual spending for each Iowa County for each year between 2010 and 2024. The spending is split based on type of spending for which the money was used.

5.  [County Population](https://data.iowa.gov/Community-Demographics/County-Population-in-Iowa-by-Year/qtnr-zsrc/about_data)

Included is the number of people in each Iowa County for each year between 1990 and 2024.

### Data Preparation and Cleaning

The first step in this process is to read the data sets using readr's read_csv function. 

```{r}
library(tidyverse)
library(lubridate)

pop_raw <- read_csv("data/County_Population_in_Iowa_by_Year_20251123.csv")

# view(pop)

income_raw <- read_csv("data/Annual_Personal_Income_for_State_of_Iowa_by_County_20251123.csv")

# view(income)

rev_raw <- read_csv("data/County_Actual_Revenues_By_Type_By_Fiscal_Year_20251123.csv")

# view(rev)

budget_raw <- read_csv("data/County_Budgeted_Expenditures_By_Service_Area_By_Fiscal_Year_20251123.csv")

# view(budget)

spend_raw <- read_csv("data/County_Actual_Expenditures_by_Service_Area_by_Fiscal_Year_20251123.csv")

# view(spend)

```

The next step is to begin cleaning each data set on various thing:
1. Selecting the correct attributes as the proper variable type
2. Making a cohesive date and county name attribute to make joining tables easier
3. Filter out data with a year not in a shared range. In this case, there is sufficient data in each data set to use the years 2010 to 2024.
4. Editing any variables to make them look nice and creating any calculated fields that might be important later.

*Population Data*

In cleaning the Iowa County Population data, the only requirement was reformatting the county name as just the name of the county in all uppercase (for example, "Story County" becomes "STORY") as well as extracting the year from the date, filtering by years between 2010 and 2024 and only keeping County, Year, and Population columns.

```{r}

pop <- pop_raw |>
  mutate(
    County = toupper(substring(County, 1, nchar(County) - 7)),
    Year = year(mdy(Year))
  ) |>
  filter(Year >= 2010, Year <= 2024) |>
  select(County, Year, Population) |>
  str()

```

*Income Data*

In cleaning the Iowa County Annual Personal Income Data, the County names had to be reformatted in uppercase and the year needed to be extracted from the date. The data had to be filtered between the years 2010 and 2024. Finally, the pivot_wider function was utilized to make distinct columns for "Personal Income" and "Personal Income Per Capita" and because the values utilized a dollar sign, the parse_number function was utilized to make it numerical and remove the dollar sign. Finally, as the personal income was in thousands of dollars, it was multiplied by 1000 to make it consistent and easily used.

```{r}

income_1 <- income_raw |>
  mutate(
    County = toupper(Name),
    Year = year(mdy(Date))
  ) |>
  filter(Year >= 2010, Year <= 2024) |>
  select(County, Year, Variable, Value)

income <- income_1 |>
  pivot_wider(names_from = Variable,
              values_from = Value
              ) |>
  rename('Personal_Income' = `Personal income`,
         'Personal_Income_Per_Capita' = `Per capita personal income`
         ) |>
  mutate(
    Personal_Income = parse_number(Personal_Income),
    Personal_Income_Per_Capita = parse_number(Personal_Income_Per_Capita),
    Personal_Income = Personal_Income * 1000
  ) |>
  str()

```

*Revenue Data*

In cleaning the Iowa County Actual Revenue Data, the County name was standardized to fit the other data sets to be all uppercase letters. The important variable here to keep was "Total Revenue" and simple percentages of how much of the County's total revenue came from what.

"Revenue_Percent_Property_Tax" is the percent of Total Revenue that comes from Property Taxes (in addition to the "Delinquent" or late property taxes either paid or failed to be paid during the year).

"Revenue_Percent_Grant" is the percent of Total Revenue that comes from "Intergovernmental" sources, most commonly as grants from the State of Iowa or the Federal Government.

"Revenue_Percent_Activities" is the percent of Total Revenue that comes from people choosing to spend money for Government goods and services such as rent, interest from borrowed money, licenses, and permits. 

"Revenue_Percent_Other" is simply the percent of Total Revenue that comes from any other source.

```{r}

rev <- rev_raw |>
  rename(
    Year = `FISCAL YEAR`,
    County = `COUNTY NAME`,
    Total_Revenue = `TOTAL REVENUES & OTHER SOURCES`
  ) |>
  mutate(
    County = toupper(County),
    Revenue_Percent_Property_Tax = round((`NET CURRENT PROPERTY TAXES` + `DELINQUENT PROPERTY TAX REVENUE`) / Total_Revenue, 4),
    Revenue_Percent_Grant = round(INTERGOVERNMENTAL / Total_Revenue, 4),
    Revenue_Percent_Activities = round((`LICENSES & PERMITS` + `CHARGES FOR SERVICE` + `USE OF MONEY & PROPERTY`) / Total_Revenue, 4),
    Revenue_Percent_Other = round(1 - Revenue_Percent_Property_Tax - Revenue_Percent_Grant - Revenue_Percent_Activities, 4)
  ) |>
  filter(Year >= 2010, Year <= 2024) |>
  select(County, Year, Total_Revenue, Revenue_Percent_Property_Tax, Revenue_Percent_Grant, Revenue_Percent_Activities, Revenue_Percent_Other) |>
  str()

```

### Data Structure

## Results

## Conclusion
